---
title: JCEM Letter - Are the Different Diabetes Subgroups Correlated With All-Cause,
  Cancer-Related, and Cardiovascular-Related Mortality?
author: Omar Yaxmehen Bello-Chavolla, Neftalí Eduardo Antonio-Villa, Luisa Fernández-Chirino,
  Arsenio Vargas-Vázquez, Jessica Paola Bahena-López
date: "23/11/2020"
output:
  html_document:
    highlight: textmate
    keep_md: yes
    keep_tex: yes
    pdf_document: null
    theme: flatly
    toc: yes
    toc_depth: '3'
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(SAScii); library(tidyverse); library(survival); library(ggpubr); library(ggsci)
nhanes3<-read.csv("nhanes3.csv")
nhanes3<- nhanes3%>% filter(!is.na(GAPGAD65) & !is.na(C1PSI) & HSAGEIR.x>=20)
nhanes3$diabetes<-as.numeric((nhanes3$GHP>=6.5 |nhanes3$G1P>=126 | nhanes3$HAD1==1))
nhanes3fin<- nhanes3 %>% filter(diabetes==1 & (C1PSI!=88888 &  G1PSI!=888888 & GHP!=8888 & BMPBMI!=8888))
cox.imp<-function(x){
  var<-summary(pool(x))[1]
  HR<-exp(summary(pool(x))[2])
  lwr<-exp(summary(pool(x))[2]-1.96*summary(pool(x))[3])
  upr<-exp(summary(pool(x))[2]+1.96*summary(pool(x))[3])
  pval<-summary(pool(x))[6]
  print(cbind(var,HR, lwr, upr, pval))
}
```

## Methods

### Data extraction
We extracted data from NHANES-III, estimating HOMA2-IR and HOMA2-B using fasting glucose concentrations and C-peptide with the HOMA2-IR calculator from Oxford. We evaluated cases with diabetes defined as previous diagnosis of diabetes HbA1c >6.5% or fasting glucose >126mg/dL, as recommended by ADA. 

```{r eval=FALSE}
library(SAScii); library(tidyverse); library(survival); library(ggpubr)

#### Download NHANES-III ####  
download.file("https://wwwn.cdc.gov/nchs/data/nhanes3/24a/ssgad.xpt", tf0 <- tempfile(), mode="wb")
GAD65 <- foreign::read.xport(tf0)
#### Household data ###
download.file("https://wwwn.cdc.gov/nchs/data/nhanes3/1a/adult.dat", tf <- tempfile(), mode="wb")
nhanes3<-read.SAScii(fn = tf,sas_ri = "https://wwwn.cdc.gov/nchs/data/nhanes3/1a/adult.sas")
write.csv(nhanes3, "nhanes3.csv")

#### Laboratory data ###
download.file("https://wwwn.cdc.gov/nchs/data/nhanes3/1a/lab.dat", tf1 <- tempfile(), mode="wb")
nhanes3_lab<-read.SAScii(fn = tf1,sas_ri = "https://wwwn.cdc.gov/nchs/data/nhanes3/1a/lab.sas")
write.csv(nhanes3_lab, "nhanes3_lab.csv")
#### Examination data ###
download.file("https://wwwn.cdc.gov/nchs/data/nhanes3/1a/exam.dat", tf1 <- tempfile(), mode="wb")
nhanes3_exam<-read.SAScii(fn = tf1,sas_ri = "https://wwwn.cdc.gov/nchs/data/nhanes3/1a/exam.sas")

#### Merge datasets ###
library(tidyverse)
nhanes3_data<- nhanes3_lab %>% left_join(GAD65, by="SEQN")
nhanes3_data<-nhanes3_data %>% left_join(nhanes3, by="SEQN")
nhanes3_data<-nhanes3 %>% left_join(nhanes3_exam, by="SEQN")

write.csv(nhanes3_data, "nhanes3.csv")

#### Dataset clusters ####
nhanes3<-read.csv("nhanes3.csv")
nhanes3<- nhanes3%>% filter(!is.na(GAPGAD65) & !is.na(C1PSI) & HSAGEIR.x>=20)
nhanes3$diabetes<-as.numeric((nhanes3$GHP>=6.5 |nhanes3$G1P>=126 | nhanes3$HAD1==1))
nhanes3fin<- nhanes3 %>% filter(diabetes==1 & (C1PSI!=88888 &  G1PSI!=888888 & GHP!=8888 & BMPBMI!=8888))
clusters<-nhanes3fin %>% dplyr::select(SEQN,GHP, HAD5R, G1PSI, C1PSI,BMPBMI, HSAGEIR.x,GAPGAD65)
names(clusters)<-c("SEQN", "HbA1c", "ADD", "GLU", "CPEP", "BMI", "AGE", "GAD65")
clusters$ADD[is.na(clusters$ADD) | clusters$ADD==999 | clusters$ADD==888]<-clusters$AGE
cox.imp<-function(x){
  var<-summary(pool(x))[1]
  HR<-exp(summary(pool(x))[2])
  lwr<-exp(summary(pool(x))[2]-1.96*summary(pool(x))[3])
  upr<-exp(summary(pool(x))[2]+1.96*summary(pool(x))[3])
  pval<-summary(pool(x))[6]
  print(cbind(var,HR, lwr, upr, pval))
}
```

### Diabetes cluster classification

We used HbA1c, body-mass index (BMI), age at diabetes diagnosis, HOMA2-IR and HOMA2-B to clasify diabetes clusters using the SNNN algorithm previously developed by our team (https://uiem.shinyapps.io/diabetes_clusters_app/). Once we classified these subgroups, we classified severe autoimmune diabetes (SAID) using anti-GAD65 antibodies with values >0.069 indicating positivity. 


```{r}
clust<-read.csv("clusters.csv")
clust0<-clust%>%select(SEQN, class, HOMA2_B, HOMA2_IR)
nhanes<-nhanes3fin %>% left_join(clust0, by="SEQN")
nhanes$Cluster<-as.character(nhanes$class)
nhanes$Cluster[nhanes$GAPGAD65>=0.069]<-"said"
table(nhanes$Cluster, nhanes$class)
nhanes$HAD5R[is.na(nhanes$HAD5R) | nhanes$HAD5R==999 | nhanes$HAD5R==888]<-nhanes$HSAGEIR.x

nhanes$years<-nhanes$HSAGEIR.x-nhanes$HAD5R
nhanes$recent<-ifelse(nhanes$years<5,1,0)
nhanes$recent<-factor(nhanes$recent, labels = c("<5 years since diagnosis", ">=5 years since diagnosis"))
mort<-read.csv("nhanes3_mort.csv")
mort$SEQN<-mort$seqn
nhanes0<-nhanes %>% left_join(mort, by="SEQN")
nhanes0$cluster2<-factor(nhanes0$Cluster, labels = c("MARD", "MOD", "SAID", "SIID", "SIRD"))
nhanes0$cluster2<-relevel(nhanes0$cluster2, ref="MOD")

## Evaluating diabetes frequencies based on years since diabetes diagnosis
t1<-table(nhanes0$recent, nhanes0$cluster2)
knitr::kable(t1, align = "lccrr")

### Plotting cluster frequencies
df <- nhanes0 %>%
  group_by(cluster2, recent) %>%
  summarise(Counts = n()) %>%
  mutate(freq = Counts / sum(Counts)) %>% drop_na()
df<-as.data.frame(df)
df$Counts<-as.numeric(df$Counts);df$freq<-as.numeric(df$freq)


ggplot(data=df, aes(y=freq, x=cluster2,fill=cluster2)) +
  geom_bar(stat="identity", position='dodge')+scale_fill_nejm()+
  scale_y_continuous(labels = scales::percent_format())+labs(fill="Cluster")+
  theme_pubr()+ylab("Frequency (%)")+xlab("")+facet_wrap(~recent)+
  theme(axis.text.x = element_blank(),axis.ticks.x = element_blank())+
  geom_text(aes(label=round(freq*100,1)), vjust=1.6, color="white", size=3.5)
```

We identified that the proportion of diabetes subgroups was modified by defining recent vs. non-recent diabetes diagnosis. The profiles of each biomarker are consistent with what is expected for each cluster.

```{r warning=FALSE}
g1<-ggplot(nhanes0, aes(x=cluster2, y=HOMA2_IR, fill=cluster2))+geom_boxplot()+scale_y_log10()+
  ylab("HOMA2-IR")+xlab("Cluster")+theme_classic()+labs(fill="Cluster")+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
g2<-ggplot(nhanes0, aes(x=cluster2, y=HOMA2_B, fill=cluster2))+geom_boxplot()+scale_y_log10()+
  ylab("HOMA2-B")+xlab("Cluster")+theme_classic()+labs(fill="Cluster")+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
g3<-ggplot(nhanes0, aes(x=cluster2, y=GHP, fill=cluster2))+geom_boxplot()+ylab("HbA1c (%)")+
  xlab("Cluster")+theme_classic()+labs(fill="Cluster")+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
g4<-ggplot(nhanes0, aes(x=cluster2, y=G1P, fill=cluster2))+geom_boxplot()+scale_y_log10()+
  ylab("Glucose (mg/dL)")+xlab("Cluster")+theme_classic()+labs(fill="Cluster")+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
g5<-ggplot(nhanes0, aes(x=cluster2, y=BMPBMI, fill=cluster2))+geom_boxplot()+scale_y_log10()+
  ylab("BMI (Kg/m2)")+xlab("Cluster")+theme_classic()+labs(fill="Cluster")+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
g6<-ggplot(nhanes0, aes(x=cluster2, y=HAD5R, fill=cluster2))+geom_boxplot()+ylab("Age at diagnosis (years)")+
  xlab("Cluster")+theme_classic()+labs(fill="Cluster")+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

ggarrange(g1, g2, g3, g4, g5, g6, labels=LETTERS[1:6])

```

### All-cause mortality

Next, we fitted a Cox proportional risk regression model to predict all-cause mortality. Time of follow-up was estimated from examination until last record. We performed multiple imputation by chained equations using the `mice` R packag with 5 iterations and five multiply imputed datasets. 

```{r, results='hide', message=FALSE}
nhanes0$mort_cvd_rec<-NULL;
nhanes0$mort_cvd_rec[nhanes0$ucod_leading==1]<-1;
nhanes0$mort_cvd_rec[nhanes0$ucod_leading>1]<-2;
nhanes0$mort_cvd_rec[nhanes0$ucod_leading==0]<-0
nhanes0$mort_cvd_rec<-na.tools::na.replace(nhanes0$mort_cvd_rec,0)

nhanes0$mort_db_rec<-NULL;
nhanes0$mort_db_rec[nhanes0$ucod_leading==2]<-1;
nhanes0$mort_db_rec[nhanes0$ucod_leading!=2]<-2;
nhanes0$mort_db_rec[nhanes0$ucod_leading==0]<-0
nhanes0$mort_db_rec<-na.tools::na.replace(nhanes0$mort_db_rec,0)



nhanes2<-nhanes0 %>% select(cluster2, mortstat, permth_exm, HSSEX.x,DMARACER.x,
                            HAR3,HAD6,HAD10,HAE5A,HSAGEIR.x,HAE8D,mort_cvd_rec,mort_db_rec, years)
nhanes2<-nhanes2[!is.na(nhanes2$mortstat) & !is.na(nhanes2$cluster2),]
nhanes3<-nhanes2 %>%filter(years<=5)

set.seed(123);imp<-mice::mice(nhanes2, maxit=5, m=5)
set.seed(123);imp2<-mice::mice(nhanes3, maxit=5, m=5)
```

To account for multiple imputation, we used R pooling function to obtain estimates of all-cause mortality for each cluster, using mild-obesity related diabetes (MOD) as the reference category.

```{r, echo=FALSE}
#### All cause mortality ####
library(mice)

### Unadjusted

m1<-with(imp, coxph(Surv(permth_exm, mortstat)~cluster2))
summary(pool(m1))
cox.imp(m1)

### Adjusted by sex, race/ethnicity and medications
m1<-with(imp,coxph(Surv(permth_exm, mortstat)~cluster2+HSSEX.x+DMARACER.x+HAR3+HAD6+HAD10+HAE5A+HAE8D))
summary(pool(m1))
cox.imp(m1)

### Adjusted by sex, race/ethnicity, medications and age
m1<-with(imp,coxph(Surv(permth_exm, mortstat)~cluster2+HSSEX.x+DMARACER.x+HAR3+HAD6+HAD10+HAE5A+HAE8D+HSAGEIR.x))
summary(pool(m1))
cox.imp(m1)

```

### Cardiovascular and cancer related mortality

To estimate cause-specific mortality, we fitted semi-parametric Finn & Gray models to account for competing risks using the `survival` R package. We first present results of cardiovascular mortality.

#### Cardiovascular mortality

```{r}
nhanes2<-nhanes0 %>% select(cluster2, mortstat, permth_exm, HSSEX.x,DMARACER.x,
                            HAR3,HAD6,HAD10,HAE5A,HSAGEIR.x,HAE8D,mort_cvd_rec,mort_db_rec,ucod_leading)
nhanes2<-nhanes2[!is.na(nhanes2$mortstat) & !is.na(nhanes2$cluster2) & !is.na(nhanes2$ucod_leading),]
set.seed(123);imp<-mice::mice(nhanes2, maxit=1, m=1)
nhanes3<-complete(imp, "long")

nhanes3$mort_cvd_rec<-NULL;
nhanes3$mort_cvd_rec[nhanes3$ucod_leading==1]<-1;
nhanes3$mort_cvd_rec[nhanes3$ucod_leading>1]<-2;
nhanes3$mort_cvd_rec[nhanes3$ucod_leading==0]<-0
nhanes3$mort_cvd_rec<-na.tools::na.replace(nhanes3$mort_cvd_rec,0)
nhanes3$cluster2<-factor(nhanes3$cluster2,ordered = F)
nhanes3$cluster2<-relevel(nhanes3$cluster2, ref="MOD")
event<- nhanes3$mort_cvd_rec
event<- factor(event, 0:2, labels=c("censor","CVD","death_other"))

### Unadjusted
pdata_2 <- finegray(Surv(permth_exm, event) ~ cluster2, data=nhanes3, na.action=na.omit)
fgfit <- coxph(Surv(fgstart, fgstop, fgstatus) ~ cluster2,weight=fgwt, data=pdata_2)
summary(fgfit)

### Adjusted
pdata_2 <- finegray(Surv(permth_exm, event) ~ cluster2+HSSEX.x+DMARACER.x+HAR3+HAD6+HAD10+HAE5A+HSAGEIR.x+HAE8D, data=nhanes3, na.action=na.omit)
fgfit <- coxph(Surv(fgstart, fgstop, fgstatus) ~ cluster2+HSSEX.x+DMARACER.x+HAD6+HAD10+HAE5A+HSAGEIR.x+HAE8D,weight=fgwt, data=pdata_2)
summary(fgfit)

```

#### Cancer-related mortality

```{r}
nhanes0$mort_ca_rec<-NULL;
nhanes0$mort_ca_rec[nhanes0$ucod_leading==2]<-1;
nhanes0$mort_ca_rec[nhanes0$ucod_leading!=2]<-2;
nhanes0$mort_ca_rec[nhanes0$ucod_leading==0]<-0
nhanes0$mort_ca_rec<-na.tools::na.replace(nhanes0$mort_ca_rec,0)

table(nhanes0$cluster2, nhanes0$mort_ca_rec)

event<- nhanes0$mort_ca_rec
event<- factor(event, 0:2, labels=c("censor","Cancer","death_other"))

### Adjusted
pdata_2 <- finegray(Surv(permth_exm, event) ~ cluster2, data=nhanes0, na.action=na.omit)
fgfit <- coxph(Surv(fgstart, fgstop, fgstatus) ~ cluster2,weight=fgwt, data=pdata_2)
summary(fgfit)

### Adjusted
pdata_2 <- finegray(Surv(permth_exm, event) ~ cluster2+HSSEX.x+DMARACER.x+HAR3+HAD6+HAD10+HAE5A+HSAGEIR.x+HAE8D, data=nhanes0, na.action=na.omit)
fgfit <- coxph(Surv(fgstart, fgstop, fgstatus) ~ cluster2+HSSEX.x+DMARACER.x+HAD6+HAD10+HAE5A+HSAGEIR.x+HAE8D,weight=fgwt, data=pdata_2)
summary(fgfit)
```


